---
title: "Scenario analysis for the Noronha trophic network"
output: html_document
---

# Scenario analysis

This document extends the original analysis by adding two complementary analyses:

1. **Rat diet uncertainty** (sampling uncertainty in the PSIRI estimates).
2. **Management scenarios** (changes in invasive predator abundance).

```{r setup}
library(data.table)

# Objects assumed to exist from the main workflow:
# psiri
# rawFreq
# spp
# samplesInvaders
# densities
# R.vec
# indirect()
```

# Helper function

```{r}
build_M <- function(psiri, rawFreq, samplesInvaders, densities){

  Bw <- rawFreq

  Bw[, names(samplesInvaders)] <-
    Bw[, names(samplesInvaders)] /
    matrix(samplesInvaders,
           nrow = nrow(Bw),
           ncol = length(samplesInvaders),
           byrow = TRUE)

  Bw[, names(samplesInvaders)] <-
    Bw[, names(samplesInvaders)] *
    matrix(densities,
           nrow = nrow(Bw),
           ncol = length(densities),
           byrow = TRUE)

  0.1 * t(psiri) + Bw
}
```

# 1. Rat diet uncertainty

The rat diet is based on only 10 samples. We generate plausible alternative
PSIRI vectors using a Dirichlet approximation centred on the observed diet.

```{r}
# install.packages("gtools")
library(gtools)

simulate_rat_psiri <- function(psiri,
                               rat = "rattus_rattus",
                               n = 10){

  p <- psiri[, rat]

  p[p < 0] <- 0
  p <- p / sum(p)

  alpha <- p * n + 1e-6

  psiri_new <- psiri
  psiri_new[, rat] <- as.numeric(rdirichlet(1, alpha))

  psiri_new
}
```

Run 1000 uncertainty realizations.

```{r}
set.seed(123)

n_sim <- 1000

rat_results <- vector("list", n_sim)

for(i in seq_len(n_sim)){

  psiri_i <- simulate_rat_psiri(psiri)

  M_i <- build_M(
    psiri = psiri_i,
    rawFreq = rawFreq,
    samplesInvaders = samplesInvaders,
    densities = densities
  )

  rat_results[[i]] <- indirect(M_i, R.vec)$Tmat
}
```

# 2. Management scenarios

Each value represents the remaining fraction of the original abundance.

```{r}
management <- list(

  baseline = c(
    felis_catus = 1,
    rattus_rattus = 1,
    salvator_merianae = 1,
    rhinella = 1
  ),

  cat_control = c(
    felis_catus = 0.1,
    rattus_rattus = 1,
    salvator_merianae = 1,
    rhinella = 1
  ),

  rat_control = c(
    felis_catus = 1,
    rattus_rattus = 0.1,
    salvator_merianae = 1,
    rhinella = 1
  ),

  tegu_control = c(
    felis_catus = 1,
    rattus_rattus = 1,
    salvator_merianae = 0.1,
    rhinella = 1
  ),

  toad_control = c(
    felis_catus = 1,
    rattus_rattus = 1,
    salvator_merianae = 1,
    rhinella = 0.1
  ),

  cat_rat = c(
    felis_catus = 0.1,
    rattus_rattus = 0.1,
    salvator_merianae = 1,
    rhinella = 1
  ),

  all_invaders = c(
    felis_catus = 0.1,
    rattus_rattus = 0.1,
    salvator_merianae = 0.1,
    rhinella = 0.1
  )
)
```

```{r}
run_management <- function(psiri,
                           rawFreq,
                           samplesInvaders,
                           densities,
                           reductions){

  densities_new <- densities

  densities_new[names(reductions)] <-
    densities_new[names(reductions)] * reductions

  M <- build_M(
    psiri = psiri,
    rawFreq = rawFreq,
    samplesInvaders = samplesInvaders,
    densities = densities_new
  )

  indirect(M, R.vec)
}
```

Run all scenarios.

```{r}
scenario_results <-
  lapply(management, function(x)
    run_management(
      psiri = psiri,
      rawFreq = rawFreq,
      samplesInvaders = samplesInvaders,
      densities = densities,
      reductions = x
    )
  )
```

# 3. Combined uncertainty + management

```{r}
combined_results <- list()

set.seed(123)

for(sim in seq_len(200)){

  psiri_i <- simulate_rat_psiri(psiri)

  for(sc in names(management)){

    combined_results[[paste(sim, sc, sep = "_")]] <-
      run_management(
        psiri = psiri_i,
        rawFreq = rawFreq,
        samplesInvaders = samplesInvaders,
        densities = densities,
        reductions = management[[sc]]
      )
  }
}
```

